### Correlation Matrix

#### Load the Dataset
```{r}
data <- readRDS("diabetic_cleaned.rds")
summary(data)
```

#### Variable Selection
In correlation analysis, only numeric variables are applicable. Therefore, the steps below outline the selection of numerical variables. 
```{r}
# Select only numeric columns
numeric_data <- data[sapply(data, is.numeric)]

# Check selected columns
str(numeric_data)
```

#### Normality Assumption
The normality assumption needs to be assessed to determine the appropriate type of correlation analysis—-whether to use Pearson or Spearman. If the data are normally distributed, Pearson will be used; otherwise, Spearman correlation will be applied.
```{r}
# Set up plotting layout
par(mfrow = c(2, 2))  # 2x2 grid for each variable

# Loop through columns
for (col in names(numeric_data)) {
  # Skip if no variance or too many NAs
  if (sd(numeric_data[[col]], na.rm = TRUE) == 0 || sum(is.na(numeric_data[[col]])) > 0.5 * nrow(numeric_data)) next
  
  # Histogram with density curve
  hist(numeric_data[[col]], 
       breaks = 30, 
       prob = TRUE, 
       main = paste("Histogram of", col),
       xlab = col)
  lines(density(numeric_data[[col]], na.rm = TRUE), col = "red", lwd = 2)
  
  # Q-Q plot
  qqnorm(numeric_data[[col]], main = paste("Q-Q Plot of", col))
  qqline(numeric_data[[col]], col = "red")
}

# Reset plotting layout
par(mfrow = c(1, 1))
```

***CONCLUSION:***
The histogram shows that most of the independent variables are not normally distributed. Therefore, Spearman correlation analysis will be used to assess the relationships. 

#### Correlation Matrix
Next, the correlation matrix will be calculated. If any variable has a standard deviation of zero (0), it will be removed from the list of numerical variables, and the matrix will be recalculated.
```{r}
# Calculate the correlation matrix [Output: warning]
cor_matrix <- cor(numeric_data, use = "complete.obs", method = "spearman")

# Find columns with zero standard deviation
zero_sd_col <- names(numeric_data)[sapply(numeric_data, function(x) sd(x, na.rm = TRUE) == 0)]
print(zero_sd_col)

# Remove columns with zero standard deviation
numeric_data_filtered <- numeric_data[, sapply(numeric_data, function(x) sd(x, na.rm = TRUE) != 0)]

# Recompute correlation matrix
cor_matrix <- cor(numeric_data_filtered, use = "complete.obs", method = "spearman")
print(round(cor_matrix, digits = 4))
```

#### Plotting the Correlation Matrix
To facilitate interpretation, a heatmap of the correlation matrix is presented below.
```{r}
# Load required packages
library(corrplot)

# Plot the correlation matrix
corrplot(cor_matrix, method = "color", addCoef.col = "black", number.cex = 0.7, tl.cex = 0.8)
```

***CONCLUSION:***
This correlation matrix shows relationships between healthcare-related variables, with values ranging from -0.70 (strong negative) to 1.00 (perfect positive). Most correlations are weak, but a few stand out as moderately strong.

**Strongest Negative Correlation**

**i)** The **strongest relationship** is between *admission_type_id* and *admission_source_id* (**-0.70**), indicating these variables move in opposite directions.  
📌 For example, emergency admissions might come from different sources than planned admissions.

**Hospital Stay Duration Relationships**

**i)** *time_in_hospital* **moderately correlates** with:  
&nbsp;&nbsp;&nbsp;&nbsp;a. *num_medications* (**0.47**) — patients staying longer tend to receive more medications.  
&nbsp;&nbsp;&nbsp;&nbsp;b. *num_lab_procedures* (**0.34**) — longer stays involve more lab tests.  

📌 Longer hospitalization is associated with increased medication and testing.

**Treatment Intensity Interconnections**

**i)** *num_procedures* and *num_medications* (**0.35**) — patients undergoing more procedures also receive more medications.  

**ii)** *number_diagnoses* and *num_medications* (**0.29**) — patients with more diagnoses tend to receive more medications.  

📌 These moderate correlations suggest that treatment intensity variables reinforce each other.

**Artificial Correlation Between Identifiers**

**i)** *encounter_id* and *patient_nbr* show a correlation of **0.54**, but this is artificial.  
📌 These are identifiers and not clinical features — their correlation reflects ID generation patterns, not medical trends.


#### Regrouping Data
The initial correlation analysis across all observations showed limited correlation between variables. However, medical data often contains patient-specific patterns that can be obscured in population-level analyses. By regrouping the data by patient number and analyzing correlations within patients, we account for between-patient variability and are better able to detect meaningful within-patient relationships between variables.
```{r}
# Load required packages
library(dplyr)

data_sum <- data %>%
  group_by(patient_nbr) %>%
  summarise(
    encounter_count = n_distinct(encounter_id),
    total_lab_procedures = sum(num_lab_procedures, na.rm = TRUE),
    total_procedures = sum(num_procedures, na.rm = TRUE),
    total_medications = sum(num_medications, na.rm = TRUE),
    total_outpatient = sum(number_outpatient, na.rm = TRUE),
    total_emergency = sum(number_emergency, na.rm = TRUE),
    total_inpatient = sum(number_inpatient, na.rm = TRUE),
    total_time_in_hospital = sum(time_in_hospital, na.rm = TRUE),
    total_num_of_diagnoses = sum(number_diagnoses, na.rm = TRUE)
  )

print(data_sum, width = Inf)
```

#### Plotting New Correlation Matrix
```{r}
# Calculate the correlation matrix
cor_matrix <- cor(data_sum, use = "complete.obs", method = "pearson")
cor_matrix

# Plot the correlation matrix with new variables
corrplot(cor_matrix, method = "color", addCoef.col = "black", number.cex = 0.7)
```

***CONCLUSION:***

**Strongest Correlations**

**i)** *encounter_count* has very strong positive correlations with:  
&nbsp;&nbsp;&nbsp;&nbsp;a. *total_num_of_diagnoses* (0.954)  
&nbsp;&nbsp;&nbsp;&nbsp;b. *total_medications* (0.859)  
&nbsp;&nbsp;&nbsp;&nbsp;c. *total_lab_procedures* (0.869)  
&nbsp;&nbsp;&nbsp;&nbsp;d. *total_time_in_hospital* (0.795)  

📌 Patients with more medical encounters tend to undergo more diagnostic tests, receive more medications, and have longer hospital stays.

**ii)** *total_num_of_diagnoses* is also strongly correlated with:  
&nbsp;&nbsp;&nbsp;&nbsp;a. *total_medications* (0.866)  
&nbsp;&nbsp;&nbsp;&nbsp;b. *total_lab_procedures* (0.850)  
&nbsp;&nbsp;&nbsp;&nbsp;c. *total_time_in_hospital* (0.793)  

📌 These strong positive correlations indicate that as the number of diagnoses increases, so do the medical interventions and hospital resources used, which aligns with clinical expectations.

**Moderate Correlations**

**i)** *total_time_in_hospital* is moderately correlated with:  
&nbsp;&nbsp;&nbsp;&nbsp;a. *total_medications* (0.812)  
&nbsp;&nbsp;&nbsp;&nbsp;b. *total_inpatient* (0.630)  

📌 Longer hospital stays involve more medical interventions.

**ii)** *total_inpatient* is moderately correlated with:  
&nbsp;&nbsp;&nbsp;&nbsp;a. *total_medications* (0.676)  
&nbsp;&nbsp;&nbsp;&nbsp;b. *total_num_of_diagnoses* (0.734)  

📌 Inpatient care is associated with higher treatment complexity.

**Weak or Negligible Correlations**

**i)** *patient_nbr* (likely a patient identifier) shows almost no correlation with other variables, as expected (all values close to 0).

**ii)** *total_outpatient* and *total_emergency* show weaker correlations with other variables (mostly < 0.4).
