## Classification: Predict Readmission Class 

 In this new process, we are going to find top features first, and then do the classification on the top features only (removing the low features) But first, we are removing the features with lower than 2 factors level.

```{r train/test}
# Load necessary libraries
library(tidyverse)
library(caret)
library(randomForest)
library(e1071)
library(xgboost)
library(ROCR)

# Load cleaned dataset
df <- readRDS("diabetic_recleaned.rds")

# Convert target variable to factor
df$readmitted <- as.factor(df$readmitted)

# Check class distribution
table(df$readmitted)

# Split into training and testing sets
set.seed(123)
train_index <- createDataPartition(df$readmitted, p = 0.8, list = FALSE)
train_data <- df[train_index, ]
test_data <- df[-train_index, ]
```

### Identify and remove low level factors
```{r }
# 01: Identify the problem columns
sapply(train_data, function(col) {
  is.factor(col) && length(unique(col)) < 2
})

names(which(sapply(train_data, function(col) is.factor(col) && length(unique(col)) < 2)))
```

```{r}
# 02: Drop problematic columns
bad_vars <- names(which(sapply(train_data, function(col) is.factor(col) && length(unique(col)) < 2)))
train_data_clean <- train_data[ , !(names(train_data) %in% bad_vars)]
```

```{r}
test_data_clean <- test_data[ , !(names(test_data) %in% bad_vars)]
```

```{r}
# double-check
bad_vars
names(train_data)
```
```{r}
remove_constant_factors <- function(df) {
  df[, !(sapply(df, function(col) is.factor(col) && length(unique(col)) < 2))]
}

train_data <- remove_constant_factors(train_data)
test_data <- remove_constant_factors(test_data)
```

```{r}
# double-check again
bad_vars
names(train_data)
names(test_data)
```
```{r}
# Drop Non-factor columns with a single unique value
#Sometimes model.matrix() can also break if a non-factor column (e.g. numeric) has zero variance.

constant_cols <- names(which(sapply(train_data, function(col) length(unique(col)) < 2)))
train_data <- train_data[, !(names(train_data) %in% constant_cols)]
test_data <- test_data[, !(names(test_data) %in% constant_cols)]
```

### Top features
```{r}
# 01. Create matrix for XGBoost
train_matrix <- model.matrix(readmitted ~ . -1, data = train_data)
train_label <- as.numeric(train_data$readmitted) - 1

# 02. Train the XGBoost model
library(xgboost)

xgb_model <- xgboost(
  data = train_matrix,
  label = train_label,
  nrounds = 50,
  objective = "multi:softprob",
  num_class = length(unique(train_label)),
  eval_metric = "mlogloss",
  verbose = 0
)

# 03. Get top features
importance <- xgb.importance(model = xgb_model)
print(head(importance, 15))

# 04. Plot the features
xgb.plot.importance(importance_matrix = importance[1:15, ])
```

### Classification Model
```{r}
# 01. Load libraries and data
library(ranger)
library(dplyr)

# 02. Select only top features (exclude patient_nbr & encounter_id)
selected_vars <- c(
  "readmitted",
  "number_inpatient",
  "discharge_disposition_id",
  "num_medications",
  "number_diagnoses",
  "number_emergency",
  "num_lab_procedures",
  "time_in_hospital",
  "number_outpatient",
  "num_procedures",
  "diabetesMed",
  "payer_code",
  "admission_source_id",
  "diag_1"
)

df_selected <- df[, selected_vars]

# 03. Remove columns with only one unique value
df_selected <- df_selected %>%
  select(where(~ length(unique(.)) > 1))

# 04. Train/test split
set.seed(123)
train_index <- createDataPartition(df_selected$readmitted, p = 0.8, list = FALSE)
train_data <- df_selected[train_index, ]
test_data  <- df_selected[-train_index, ]
```

### Option A: Random Forest (ranger)
```{r}
# 05A. Train with ranger
rf_model <- ranger(readmitted ~ ., data = train_data, importance = "impurity")

# 06A. Predict and evaluate
pred_rf <- predict(rf_model, test_data)$predictions
rf_conf <- confusionMatrix(pred_rf, test_data$readmitted)
rf_stats <- rf_conf$overall[c("Accuracy", "Kappa")]
```

```{r}
# Print RandomForest Confusion matrix
rf_conf
```

### Option B: XGBoost (with numeric matrix input)
```{r}
# 05B. Prepare XGBoost input
# Must convert factors to dummy vars
train_matrix <- model.matrix(readmitted ~ . -1, data = train_data)
train_label <- as.numeric(train_data$readmitted) - 1

test_matrix <- model.matrix(readmitted ~ . -1, data = test_data)
test_label <- as.numeric(test_data$readmitted) - 1

# 6B. Train XGBoost model
xgb_model <- xgboost(
  data = train_matrix,
  label = train_label,
  nrounds = 50,
  objective = "multi:softmax",
  num_class = length(unique(train_label)),
  eval_metric = "mlogloss",
  verbose = 0
)

# 7B. Predict and evaluate
pred_xgb <- predict(xgb_model, test_matrix)
pred_xgb <- factor(pred_xgb, labels = levels(train_data$readmitted))

xgb_conf <- confusionMatrix(pred_xgb, test_data$readmitted)
xgb_stats <- xgb_conf$overall[c("Accuracy", "Kappa")]
```

```{r}
# Print XGBoost Confusion matrix
xgb_conf
```

```{r}
# 08. Combine Metrics for Comparison

compare_df <- data.frame(
  Metric = names(rf_stats),
  RandomForest = round(as.numeric(rf_stats), 4),
  XGBoost = round(as.numeric(xgb_stats), 4)
)

# Print comparison
print(compare_df)

# Optional: Show class-wise stats
print(rf_conf$byClass)
print(xgb_conf$byClass)
```

### Results & Interpretation
(1) OVERALL COMPARISON

Both models are only moderately better than chance. Accuracy and Kappa are very similar, suggesting minimal difference in performance.

(2) CLASS-WISE PERFORMANCE

🔹 Class <30 (most underrepresented)
❌ Both models perform poorly on detecting <30 readmission (sensitivity = 4.4%)

🔹 Class >30
⚖️ Slightly better sensitivity with XGBoost, but difference is marginal.

🔹 Class NO (majority class)
✅ Both models favor this class. Slightly better balance with XGBoost.

(3) Interpretation

Both models are heavily biased toward the majority class (NO).
Extremely poor sensitivity for <30, which is the critical class for early readmission — this should be improved.
Accuracy isn’t a strong metric here; consider using macro/micro averaged F1 scores or AUC per class.
No major difference between models — you can prioritize the faster one (usually XGBoost).

### Using other metric for both models
```{r}
# 01. Step 1: Prepare Libraries

# List of required packages
required_packages <- c("MLmetrics", "yardstick", "dplyr", "pROC")

# Install any that aren't already installed
for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

library(MLmetrics)
library(yardstick)
library(dplyr)
library(pROC)
```

```{r}
# 02. Create Prediction Results Data Frame

# Use this — works fine if rf_model is trained with caret::train()
# Assume these are your actual and predicted labels
actual <- test_data$readmitted
pred_rf <- predict(rf_model, data = test_data)$predictions
rf_probs <- predict(rf_model, data = test_data, type = "response")$predictions

# Recreate the test matrix (same variables as during training!)
test_matrix <- model.matrix(readmitted ~ . -1, data = test_data)

# Predict class
pred_xgb <- predict(xgb_model, newdata = test_matrix)

# If model used "multi:softmax", output is class index (e.g., 0, 1, 2)
# So you may need to map back to class labels:
levels(actual)  # Check the actual class order
pred_xgb <- factor(pred_xgb, levels = 0:2, labels = levels(actual))
```
```{r}
# 03. Compute F1 Scores
# Create results data frames
rf_results <- data.frame(truth = actual, prediction = pred_rf)
xgb_results <- data.frame(truth = actual, prediction = pred_xgb)

# Convert columns to factors (yardstick requires factors)
rf_results$truth <- as.factor(rf_results$truth)
rf_results$prediction <- as.factor(rf_results$prediction)

xgb_results$truth <- as.factor(xgb_results$truth)
xgb_results$prediction <- as.factor(xgb_results$prediction)

# Compute F1 macro and micro for Random Forest
rf_fma <- f_meas(rf_results, truth = truth, estimate = prediction, average = "macro")
rf_fmi <- f_meas(rf_results, truth = truth, estimate = prediction, average = "micro")

# Compute F1 macro and micro for XGBoost
xg_fma <- f_meas(xgb_results, truth = truth, estimate = prediction, average = "macro")
xg_fmi <- f_meas(xgb_results, truth = truth, estimate = prediction, average = "micro")

# Combine into a data frame
f1_summary <- data.frame(
  Model = c("Random Forest", "Random Forest", "XGBoost", "XGBoost"),
  Metric = c("F1 Macro", "F1 Micro", "F1 Macro", "F1 Micro"),
  Estimate = c(rf_fma$.estimate, rf_fmi$.estimate, xg_fma$.estimate, xg_fmi$.estimate)
)

# Save to CSV
write.csv(f1_summary, "f1_score.csv", row.names = FALSE)
```

```{r}
f1_summary
```
### F1 Score Interpretation
✅ F1 Macro (treats all classes equally):
*On average, both models perform similarly across all classes. This metric treats each class equally, even if one class (e.g., "NO") dominates. A score below 0.5 suggests poor ability to balance precision and recall, especially for minority classes (<30, >30).

✅ F1 Micro (accounts for class imbalance):
*Interpretation: These values show how well the models perform across all instances, weighted by class frequency. Micro F1 is generally close to accuracy in multi-class problems. Both models are underperforming overall — correctly predicting fewer than half of the cases.

🔍 Final Analysis:
*Both models perform nearly the same in terms of F1 scores.
*F1 < 0.4 implies room for improvement, possibly due to:
 -Class imbalance (e.g., "NO" dominating)
 -Limited predictive power in selected features